---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
---

### Load libraries
```{r, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(scales)   
```


### Loading and preprocessing the data
Download, unzip, and read in csv

```{r, echo=TRUE}
if (!file.exists(paste0(getwd(), '/data/activity.csv')) )
{
  fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(fileUrl, destfile = paste0(getwd(), '/repdata%2Fdata%2Factivity.zip'), method = "curl")
  unzip("repdata%2Fdata%2Factivity.zip",exdir = "data")
}
df <- read.csv(paste0(getwd(), '/data/activity.csv'), stringsAsFactors = FALSE)
df$date <- as.Date(df$date)

str(df)
```



#### 1. Make Histogram of the total number of steps taken each day

```{r, echo=TRUE}
stepsByDay <- df %>% group_by(date) %>% summarize(steps = sum(steps, na.rm=TRUE))
head(stepsByDay)

hist(stepsByDay$steps, main="Histogram of Steps by Day", xlab= "Steps By Day", ylab="Number of Days", breaks=20, col="gray")
```

Note that we ignored NA values with na.rm=TRUE



####  2. Calculate the mean and median number of steps taken each day

The mean number of steps per day is
```{r, echo=TRUE}
mean(stepsByDay$steps)
``` 
and the median number of steps per day is
```{r, echo=TRUE}
median(stepsByDay$steps)
```

Next, we can add these values to the histogram 

```{r, echo=TRUE}
stepsByDay <- df %>% group_by(date) %>% summarize(steps = sum(steps, na.rm=TRUE))
hist(stepsByDay$steps, 
     main="Histogram of Steps by Day", 
     xlab= "Steps By Day", 
     ylab="Number of Days", 
     breaks=20, 
     col="gray")
abline(v=mean(stepsByDay$steps), lty=3, col="blue")
text(mean(stepsByDay$steps),10,labels="mean", pos=1, col="blue") 
abline(v=median(stepsByDay$steps), lty=4, col="red")
text(median(stepsByDay$steps),9,labels="median", pos=1, col="red") 

```



####  3. Make a Time Series Plot of the Average Number of Steps Taken

```{r, echo=TRUE}
ggplot(stepsByDay, aes(x=date, y=steps)) +   
  geom_line( color="steelblue") + 
  geom_point() + 
  xlab("Date") + 
  ylab("Number of Steps") + 
  scale_x_date(date_labels = "%b-%d-%Y") + 
  scale_y_continuous(labels=comma) + 
  ggtitle("Steps Taken in Day")
```

And then if we want to see the moving average line and band, we can add geom_smooth(method="loess") 

```{r, echo=TRUE}
ggplot(stepsByDay, aes(x=date, y=steps)) +   
  geom_line( color="steelblue") + 
  geom_point() + 
  geom_smooth(method="loess") +  #adds moving average 
  xlab("Date") + 
  ylab("Number of Steps") + 
  scale_x_date(date_labels = "%b-%d-%Y") + 
  scale_y_continuous(labels=comma) + 
  ggtitle("Steps Taken in Day - With Moving Average")
```



####  4. Find 5-minute interval that, on average, contains the maximum number of steps

First, summarize data by interval

```{r, echo=TRUE}
stepsByInterval <- df %>% group_by(interval) %>% summarize(steps = mean(steps, na.rm=TRUE))
head(stepsByInterval)
```

Then can find the maximum

```{r, echo=TRUE}
maxinterval <- stepsByInterval[which.max(stepsByInterval$steps), ]$interval
maxinterval 
```

Or can visualize it in plot

```{r, echo=TRUE}
ggplot(data=stepsByInterval, aes(x=interval, y=steps)) +
  geom_line() +
  xlab("5-Minute Interval") +
  ylab("Average Number of Steps Taken") + 
  scale_y_continuous(labels=comma) + 
  ggtitle("Average Steps Taken In 5 Minute Intervals")
```



### Imputing Missing Values



#### Calculate and report the total number of missing values in the dataset

The total number of rows with at least 1 NA is
```{r, echo=TRUE}
sum(!complete.cases(df))
```

The total number of NAs is
```{r, echo=TRUE}
sum(is.na(df))
```



#### Devise a strategy for filling in all of the missing values in the dataset

We will make a new variable 'steps2' that will fill in the missing values of 'steps' with the overall average for the interval

step 1: Summarize the data by interval

```{r, echo=TRUE}
 stepsByInterval <- df %>% group_by(interval) %>% summarize(avgsteps = mean(steps, na.rm=TRUE)) %>% ungroup()
 head(stepsByInterval)
```

step 2: Merge it back to original dataframe

```{r, echo=TRUE} 
df <- merge(df, stepsByInterval, by="interval", all.x = TRUE)
head(df)
```

step 3: Calculate 'steps2'

```{r, echo=TRUE} 
 df <- df %>% mutate(steps2 = ifelse(is.na(steps), avgsteps, steps))  
head(df)
```

step 4: Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r, echo=TRUE} 
df2 <- df %>% select(interval, date, steps2) %>% rename(steps = steps2)
head(df2)
```



#### Make Histogram of the total number of steps taken each day after missing values are imputed

```{r, echo=TRUE} 
 steps2ByDay <- df %>% group_by(date) %>% summarize(steps2 = sum(steps2, na.rm=TRUE))
 hist(steps2ByDay$steps2,  breaks=20, col="gray", main="Histogram of (Imputed) Steps by Day", xlab= "Steps By Day", ylab="Number of Days")
```



#### Calculate the mean and median number of steps taken each day

The mean number of steps per day is
```{r, echo=TRUE}
mean(steps2ByDay$steps2)
``` 
and the median number of steps per day is
```{r, echo=TRUE}
median(steps2ByDay$steps2)
```

Note that after imputing the NAs, the mean changed by
```{r, echo=TRUE}
mean(steps2ByDay$steps2) - mean(stepsByDay$steps) 
``` 
and the median changed by
```{r, echo=TRUE}
median(steps2ByDay$steps2) - median(stepsByDay$steps) 
``` 



### Weekdays vs Weekends



#### Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r, echo=TRUE}
df <- df %>% mutate(day_type = ifelse(weekdays(df$date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))
head(df)
```

#### Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r, echo=TRUE}
stepsByInterval <- df %>% group_by(day_type, interval) %>% summarize(steps = mean(steps, na.rm=TRUE)) %>% ungroup()
ggplot(data=stepsByInterval, aes(x=interval, y=steps), color=day_type) +
  geom_line() +
  xlab("5-Minute Interval") +
  ylab("Average Number of Steps Taken") + 
  scale_y_continuous(labels=comma) + 
  ggtitle("Average Steps Taken In 5 Minute Intervals") + facet_grid(day_type ~ .)
```


